### 2014-07-16 paper.tex
polished the conclusion part

% Results and Discussion can be combined.
\section*{Results}

\subsection*{Differential gene expression in response to MDV infection}

To extensively study gene and isoform expression, we incorporated ENSEMBL gene
models release 73 with {\em de novo} and a reference-guided transcriptome
assembly to build custom gene models.  The models, therefore, include both
ENSEMBL annotated transcripts and putative transcripts and genes.  The advantage
of using custom gene models is it allows an investigation of unannotated genes
and isoforms, which is necessary for in-depth transcriptome study of an immune
system.

Many genes were found to be differentially expressed (DE) between control and
infected chickens in lines 6 and line 7.  While the number of unique
downregulated genes in both lines was approximately equal, the number of unique
upregulated genes in line 7 was much greater compared to line 6
(Figure~\ref{degenes_venn})~\cite{lee1983ontogeny}.  Interestingly, some genes
that were differentially expressed in both lines were regulated in the opposite
direction (Table~\ref{tab:opposite}).  Among genes downregulated in line 6 but
upregulated in line 7 were {\em LL} (lung lectin) and {\em SFTPA1}, which encode
a calcium-dependent C-type lectin and a lung surfactant protein respectively.
Both molecules are important in innate
immunity~\cite{hogenkamp2008chicken,kingma2006defense}.  {\em LIMS1} is involved
in cell differentiation and proliferation and {\em PPARG} is a suppressor of
{\em NF$\kappa$B}-mediated proinflamatory response.  On the other hand, nearly
all genes upregulated in line 6 but downregulated in line 7 are involved in cell
survival such as mRNAs splicing, cell growth, and protein synthesis, except CD7
whose function is involved in T cell-B cell interaction.  This difference
suggests that even at this stage of infection in line 6, the lytic phase could
be subsided, therefore, only genes important for survival of cell are
upregulated in line 6.  In comparison, the lytic phase in line 7 may still
continue and as a result, genes involved in immune responses are still
upregulated.  In addition, type I interferon ({\em IFN-$\gamma$} and {\em
IFN-$\beta$}) as well as {\em INF-$\alpha$3} were found to be highly upregulated
in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors were not
different in line 6, but upregulated in line 7.  This could also reflect the
ongoing immune responses in line 7.

Some DE genes have been reported from previous microarray studies were also
found to be differentially expressed in this study.  For example, B6.1 (Bu-1)
gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4
d.p.i~\cite{sarson2008transcriptional}.  It was also found to be down-regulated
about 3-fold in line 7 (susceptible).  Similarly, {\em GMZA} gene reported to be
upregulated across genetically different chickens (B$^{19}$, B$^{21}$ alleles),
was also found to be highly upregulated.  In contrast, some genes that have been
reported to be highly expressed in resistant chickens were downregulated in both
lines. Those genes are {\em AMIGO2, MMP13} and {\em CLEC3B}, which were found to
be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed
% in resistant chickens: \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all
% those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line
7$_{2}$) including {\em AVD, ART1, NOS2, CXCL13L2, MX1} and {\em
SOCS1}~\cite{smith2011systems} were also found to be highly upregulated in both
line 6 and 7.  However, our results show the similar expression patterns for
{\em IL6} and {\em IL18}, which were only upregulated in line 7 in early stage
of infection (3-5 d.p.i).

In contrast, {\em IL15} has been reported non differential-expressed between
control and infected chickens in line 6$_{1}$ and
7$_{2}$~\cite{kaiser2003differential}; however, it was only upregulated in line
7 in our study.  Expression of {\em IL15} is induced by {\em TLR9}, which binds
to non-methylated CpG residues present in genomes of many DNA viruses, including
herpes simplex virus.  This cytokine auto-regulates the expression of {\em
CD40}, which is a transmembrane receptor required for activation of macrophages
by CD4 T cells.  Consequently, {\em CD40} was only upregulated in line 7 (data
not shown).

@@

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq
data~\cite{young2010method}.  Significantly perturbed pathways (FDR $< 0.1$)
from both lines that involved in immune response include TLR signaling pathway,
cytokine-cytokine receptor interaction, intestinal immune network for IgA
production, and cell adhesion molecules (CAMs)~\ref{line67_kegg}.  Some other
pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling
pathway, NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.  Although MHC I gene ({\em BF1}) was
differentially expressed in both lines, other genes involved in expressing newly
synthesized MHC I were only upregulated in line 7 suggesting that new MHC I
molecules were actively produced and expressed on cell surface.
% (the minor class I is upregulated but the major class I is downregulated)
Furthermore, gene ontology analysis of biological processes (GO:BP) shows that
categories involved in both adaptive and innate immune responses were enriched
in line 7 (supplementary material).  On the other hand, only categories involved
in innate immune responses were enriched in line 6.  In addition, enrichment of
apoptosis pathway in line 7 indicates that the programmed cell death could be
induced by CTL response to eliminate the viruses.

@@

\subsection*{Differential exon usage (DEU)}

@@

Due to those issues, it is not feasible to accurately estimate expression of
isoforms, especially when gene annotation is constructed from {\em de novo}
assembly~\cite{trapnell2013differential}.  To avoid these issues, we chose to
study exon expression instead of isoform expression.  Using MISO with
exon-centric method, only reads spanning across a few exons are used and only
exons involved in a splicing event are examined.  The expression of exon
inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used
to infer the portion of transcripts that include the exon in each
sample~\cite{Katz:2010iv}.  In this study, we investigated the three most common
alternative splicing events in vertebrates, which are skipped exons (SE), an
alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.

\subsubsection*{DEU genes in response to MDV}

Lists of DEU genes from line 6 that show difference in $\Psi$ greater than 0.20
when compared to line 7 in infected chickens are shown in
Tables~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.  Genes can be categorized roughly into
four groups based on the pattern of $\Psi$ across control and infected birds in
both lines.

@@

\section*{Discussion}

Custom gene models built from combination of gene models from {\em de novo}
assembly, reference-based assembly, and Ensembl allowed us to identify genes and
isoforms that might play an important role in resistance to MD.  Results from
gene expression analysis indicate that adaptive immune responses were active
during lytic infection in line 7.  Because only activated T cells are infected,
elicitation of adaptive immune responses could help spread the viruses by
recruiting and activating more T cells.  Therefore, active immune responses are
favorable for the virus.  On the other hand, the delay of adaptive immune
responses in line 6, in this case, could benefit the host by limiting infection
of activated T cells.

To conclude, we attempted to employ custom gene models built from RNA-Seq data
to identify underlying genetic factors contributing to MD resistance.  We
identified a number of genes that expressed different ratios of alternative
splice forms between lines 6 and 7.  Interestingly, several genes are involved
in actin cytoskeleton and cytokinesis, which have not been of great interest in
the filed of MD research.  Even though we mainly discussed the possible role of
{\em ITGB2} in MD resistance, other genes cannot be precluded and should be a
candidate for further investigation and experimental studies.  Moreover, a full
mechanism of MD resistance may be highly complex and require more data from
different stages of infection as well as a greater sequencing depth in order to
identify all genes and isoforms involved.  To enable the study of unannotated
gene and isoform expression, our approach of constructing gene models from
RNA-Seq should be iteratively used to extend the Ensembl gene models.

>>>>

% Results and Discussion can be combined.
\section*{Results and Discussion}

\subsection*{Differential expression results from our method are comparable to
previous studies}

To extensively study gene and isoform expression, we incorporated ENSEMBL gene
models release 73 with {\em de novo} and a reference-guided transcriptome
assembly to build custom gene models.  The models, therefore, include both
ENSEMBL annotated transcripts and putative genes and isoforms.  The advantage of
using custom gene models is it allows an investigation of unannotated genes and
isoforms, which is necessary for in-depth study of an immune system.

Some DE genes have been reported from previous microarray studies were also
found to be differentially expressed in this study.  For example, B6.1 (Bu-1)
gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4
d.p.i~\cite{sarson2008transcriptional}.  It was also found to be down-regulated
about 3-fold in line 7 (susceptible).  Similarly, {\em GMZA} gene reported to be
upregulated across genetically different chickens (B\textsuperscript{19},
B\textsuperscript{21} alleles),
was also found to be highly upregulated.  In contrast, some genes that have been
reported to be highly expressed in resistant chickens were downregulated in both
lines. Those genes are {\em AMIGO2, MMP13} and {\em CLEC3B}, which were found to
be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed
% in resistant chickens: \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all
% those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line
7) including {\em AVD, ART1, NOS2, CXCL13L2, MX1} and {\em
SOCS1}~\cite{smith2011systems} were also found to be highly upregulated in both
line 6 and 7.  However, our results show the similar expression patterns for
{\em IL6} and {\em IL18}, which were only upregulated in line 7 in early stage
of infection ($3-5$ d.p.i).

In contrast, {\em IL15} has been reported non differential-expressed between
control and infected chickens in line 6 and 7~\cite{kaiser2003differential};
however, it was only upregulated in line 7 in our study.  Expression of {\em
IL15} is induced by {\em TLR9}, which binds to non-methylated CpG residues
present in genomes of many DNA viruses, including herpes simplex virus.  This
cytokine auto-regulates the expression of {\em CD40}, which is a transmembrane
receptor required for activation of macrophages by CD4 T cells.  Consequently,
{\em CD40} was only upregulated in line 7 (data not shown).

\subsection*{Differential expression suggests inactive adaptive immune responses
in line 6}

Many genes were found to be differentially expressed (DE) between control and
infected chickens in lines 6 and line 7.  While the number of unique
downregulated genes in both lines was approximately equal, the number of unique
upregulated genes in line 7 was much greater compared to line 6
(Figure~\ref{degenes_venn}).  Interestingly, some genes
that were differentially expressed in both lines were regulated in the opposite
direction (Table~\ref{tab:opposite}).  Among genes downregulated in line 6 but
upregulated in line 7 were {\em LL} (lung lectin) and {\em SFTPA1}, which encode
a calcium-dependent C-type lectin and a lung surfactant protein respectively.
Both molecules are important in innate
immunity~\cite{hogenkamp2008chicken,kingma2006defense}.  {\em LIMS1} is involved
in cell differentiation and proliferation and {\em PPARG} is a suppressor of
{\em NF$\kappa$B}-mediated proinflamatory response.  On the other hand, nearly
all genes upregulated in line 6 but downregulated in line 7 are involved in cell
survival such as mRNAs splicing, cell growth, and protein synthesis, except CD7
whose function is involved in T cell-B cell interaction.  This difference
suggests that even at this stage of infection in line 6, the lytic phase could
be subsided, therefore, only genes important for survival of cell are
upregulated in line 6.  In comparison, the lytic phase in line 7 may still
continue and as a result, genes involved in immune responses are still
upregulated.  In addition, type I interferon ({\em IFN-$\gamma$} and {\em
IFN-$\beta$}) as well as {\em INF-$\alpha$3} were found to be highly upregulated
in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors were not
different in line 6, but upregulated in line 7.  This could also reflect the
ongoing immune responses in line 7.

@@

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq
data~\cite{young2010method}.  Significantly perturbed pathways (FDR $< 0.1$)
from both lines that involved in immune response include TLR signaling pathway,
cytokine-cytokine receptor interaction, intestinal immune network for IgA
production, and cell adhesion molecules (CAMs)(figure~\ref{line67_kegg}).  Some
other pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling
pathway, NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.  Although MHC I gene ({\em BF1}) was
differentially expressed in both lines, other genes involved in expressing newly
synthesized MHC I were only upregulated in line 7 suggesting that new MHC I
molecules were actively produced.  Furthermore, gene ontology analysis of
biological processes (GO:BP) shows that categories involved in both adaptive and
innate immune responses were enriched in line 7 (supplementary material).  On
the other hand, only categories involved in innate immune responses were
enriched in line 6.  In addition, enrichment of apoptosis pathway in line 7
indicates that the programmed cell death could be induced by CTL response to
eliminate the viruses.

@@

\subsection*{Genes with differential exon usage (DEU) in response to MDV
infection can be divided into four groups based on their patterns of expression}

@@

Due to those issues, it is not feasible to accurately estimate expression of
isoforms, especially when gene annotation is constructed from {\em de novo}
assembly~\cite{trapnell2013differential}.  To avoid these issues, we chose to
study exon expression instead of isoform expression.  Using MISO with
exon-centric method, only reads spanning across a few exons are used and only
exons involved in a splicing event are examined.  The expression of exon
inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used
to infer the portion of transcripts that include the exon in each
sample~\cite{Katz:2010iv}.  In this study, we investigated the three most common
alternative splicing events in vertebrates, which are skipped exons (SE), an
alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.
Lists of DEU genes from line 6 that show difference in $\Psi$ greater than 0.20
when compared to line 7 in infected chickens are shown in
Tables~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.  Genes can be categorized roughly into
four groups based on the pattern of $\Psi$ across control and infected birds in
both lines.

@@

\section*{Conclusion}

Custom gene models built from combination of gene models from {\em de novo}
assembly, reference-based assembly, and ENSEMBL had allowed us to identify genes
and isoforms that might play an important role in resistance to MD.  Results
from gene expression analysis indicated that adaptive immune responses were
active during lytic infection in line 7, but not in line 6.  Because only
activated T cells are infected by MDV, we speculated that elicitation of
adaptive immune responses could help spread the viruses by recruiting and
activating more T cells and.  In contrast, the delay of adaptive immune
responses could benefit the host by limiting infection of activated T cells.

To elucidate the molecular mechanism of MD resistance, we investigated
differential isoforms expression between lines 6 and 7 and identified a number
of genes that could be responsible for difference in immune responses. Notably,
several genes are involved in actin cytoskeleton and cytokinesis, which are
important for the functions of lymphocytes and immune cells but have not been of
great interest in the filed of MD research.  Even though we mainly discussed the
possible role of {\em ITGB2} in MD resistance, other genes cannot be precluded
and should be a candidate for further investigation and experimental studies.
Moreover, a full mechanism of MD resistance is highly complex and more data from
different stages of infection as well as a greater sequencing depth will be
required to identify all genes and isoforms involved.  To enable the study of
unannotated gene and isoform expression, our approach of constructing gene
models from RNA-Seq should be iteratively used to extend the ENSEMBL gene models
to construct more complete gene models.
